Explaining probabilistic predictions on the simplex with Shapley compositions

### 2024-01-07 paper/main.tex
correct proof read

Compositional data analysis \cite{aitchison1982,pawlowskymodeling} aims at treating data---known as compositions---living on a simplex. Compositional data analysis tools have been applied for instance on geological, and chemical data but also to discrete probability distributions \cite{egozcue2011evidence,egozcue2018evidence,noe2023representing}. This paper presents the use of the Aitchison geometry of the simplex, coming from the field of compositional data analysis, for extending the concept of Shapley value to the probability simplex, i.e. the space of discrete probability distributions. This can be used to properly explain multiclass probabilistic prediction.

@@

\section{Explaining probabilistic prediction with Shapley compositions}
\label{sec:explain}

Given a prediction $\bm{f}(\bm{x})$, the Shapley composition $\bm{\phi}_{\bm{f},\bm{x},\text{Pr}}(i)$ describes the contribution of the $i$th feature on the prediction. The efficiency property shows how the probability distribution moves from the base value, i.e. the expected prediction regardless of the current input, to the prediction $\bm{f}(\bm{x})$. In the standard Shapley formulation recalled in Section \ref{sec:shapley}, the prediction is one-dimensional such that the Shapley quantity is a scalar. In application where there are more than two possible classes, the prediction is multidimensional such that the Shapley quantity is too. Both lives on the same space: the probability simplex. In this section, we discuss how the set of Shapley compositions can be analysed to better understand the contribution and influence of each features on the prediction.

@@

The Shapley compositions can be visualized in the Euclidean space isometric to the simplex thanks to the ILR transformation presented in Section \ref{sec:ilr}. This space has the advantage of being intuitive since it is a standard real $(N-1)$-dimensional vector space.

@@

In the three classes case, the space is $2$-dimensional. We illustrate this example with the well known Iris classification dataset consisting of a set of flowers descriped by 4 features: sepal length and width and petal length and width. The aim is to predict to which of the three species, setosa, versicolor and virginica, a flower belongs to. In our example we use a Support Vector Machine (SVM) classifier with a radial basis function (rbf) kernel as a classifier. Figure \ref{fig:3classes} shows the explanation of one versicolor instance where \ref{fig:3classesshap} shows the Shapley composition in the ILR space and \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction. Having the highest norm, the petal length is the feature contributing the most on the prediction, and moves the base distribution to the versicolor maximum probability decision region (maximum probability decision region boundaries are the dashed gray lines). Being orthogonal to the virginica class-composition, this suggest that this features does not contribute on the predicted probability for this class. The Shapley composition for the petal width goes straight to the opposite direction of the setosa class vector suggesting that this feature contributes in rejecting this class. The other Shapley compositions have a low norm suggesting these features does not contribute in the prediction.
\begin{figure}

  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplot.png}
    \caption{Shapley compositions in the ILR space.}
    \label{fig:3classesshap}
  \end{subfigure}

\vspace{0.5cm}
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
    \caption{Sum of the Shapley compositions in the ILR space from the base distribution to the prediction.}
    \label{fig:3classesshapsum}
  \end{subfigure}

  \caption{Shapley explaination in the ILR space for the classification of an Iris instance.}
  \label{fig:3classes}
\end{figure}

@@

In a four classes example, the simplex is $3$-dimensional. We illustrate this with a simple digit recognition task\footnote{We use the scikit-learn's digits dataset \cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digit among: 0, 1, 2 and 3. Since they are 64 pixels, considering each pixel as a feature would correspond to 64 Shapley composition. We therefore reduce the number of features to 6 using a principal component analysis for better clarity and conciseness. We again use a SVM classifier with a rbf kernel. The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated by Figure \ref{fig:4classesshapsum}. To better understand how this space is divided into four regions each representing the maximum probabity region for one class, one can think about the shape of a methane molecule. The hydrogens correspond to the vextices and the carbon to the center of a tetrahedron i.e. a $3$-dimensional simplex. The relative position of the class-compositions in the ILR space are the same as the bonds between the carbon and a hydrogen: the angles are $\approx 109.5^{\circ}$. In this example, the tested instance is a 0\footnote{More examples can be obtained from the notebooks released on the git repository: link not shared during the double blind peer reviewing.}.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction. The gray transparent walls markout the maximum probability decision regions.}
  \label{fig:4classesshapsum}
\end{figure}

@@

When more than three classes are involved, all the dimensions of the ILR space cannot be visualized at once, but $2$ or $3$-dimensional subspaces can still be visualized. In order to select the ILR components to visualize, one needs to understand what they refer to. In this section, we briefly discuss the interpretation of the ILR components.% of the explanation of a classifier prediction with more than three classes.

A component of the ILR space  can be interpreted as a \emph{balance}, i.e. a log-ratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}: one giving the central values of the probabilities in one group of classes and one for another group of classes. Therefore, a balance is here comparing the weight of two groups of classes. The set of balances is built such that they are geometrically orthogonal meaning they provide nonredundant information\footnote{Not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}.

This can be illustrated by a sequential binary partition or bifurcation tree. Figure \ref{fig:trees} gives two examples: \ref{fig:bifurc1} shows the bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$. The first balance $\tilde{p}_1$ first compare the probabilities of class $1$ and $2$. Each next balance then recursively compares the probability for the next class with the probabilities for the previous ones independently of all the others.

In some applications, one may be interested in particular comparisons of groups of classes. For instance, like in an example presented in \cite{egozcue2005groups}, if one wants to compare political parties or groups, it may be pertinent to have a balance comparing left and right-wing groups. However, there are sometimes no obvious relevant comparisons to study. For instance, in the handwritten digit recognition problem, it may first seem natural to compare odd with even numbers or prime with non-prime but being basically a shape recognition problem, and the shape of the numbers being independent of their arithmetic properties, these comparisons are not pertinent. Be that as it may, the choice of the basis must be left open, whether or not it is based on a relevant strategy. Let's use the basis of Figure \ref{fig:bifurc2} for a 10-classes digit recognition classification\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging pair of classes based on the Mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with same covariance matrix.}. We comment, for conciseness, a single $2$-dimensional subspace\footnote{More visualisations can be obtaned from the shared python notebooks.}. Let's have a look, in Figure \ref{fig:moreclasses35}, at the third and fifth ILR dimensions ($\tilde{p}_3$ and $\tilde{p}_5$). It is like saying we are only interested in comparing the output probability for class 0 and for class 6, and in comparing the probability for class 1 with the group of probabilities for classes 7 and 8. $\tilde{p}_3$ depends only on the probability for the digits 0 and 6 and $\tilde{p}_5$ depends only on the probabilities for the digits 1, 7 and 8. Therefore, the class-composition for the others digits have a zero projection within this subspace and are not drawn in Figure \ref{fig:moreclasses35}. The projection of the class-compositions for 0 and 6 are orthogonal to the ones for class 1, 7 and 8. Indeed, the set of classes making the balance $\tilde{p}_3$ and the set of classes making $\tilde{p}_5$ have no intersection contrary to the example of Figure \ref{fig:3classes} where $\tilde{p}_1$ is comparing the probabilities for the class setosa with the probability for the class versicolor and $\tilde{p}_2$ is comparing the probabilities for the class virginica with the group of probabilities for setosa and versicolor. In this case, none of the class-compositions are orthogonal. In Figure \ref{fig:moreclasses35}, since $\tilde{p}_5$ is comparing 1 with the group of digits 7 and 8, the projection of the class-compositions on this line for 1 goes in an opposite direction than the one for the class-compositions for 7 and 8. The two latters, are equal and are half as long as the former. In this way, $\tilde{p}_5$ compares the probability for 1 with the group of probabilities for 7 and 8 with the same weight.

@@

  \vspace{0.5cm}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \input{bifurctree1.tex}
      \caption{Bifurcation tree used in our $10$-classes digit recognition task}
      \label{fig:bifurc2}
    \end{subfigure}
  \caption{Two examples of bifurcation tree.}
  \label{fig:trees}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{figures/moreclasses/ilrplot35.png}
  \caption{Sum of the Shapley compositions from the base to the prediction in the ILR subspace made of $\tilde{p}_3$ and $\tilde{p}_5$ for a test instance from the class 2. $\tilde{p}_3$ compares the probability for the class 0 with the one for the class 6 and $\tilde{p}_5$ compares the probability for the class 1 with the group of probabilities for the class 7 and 8. }
  \label{fig:moreclasses35}
\end{figure}
\subsection{Angles, norms and projections}

@@

The use of standard Shapley value for explaining machine learning model with multidimensional probabilistic output has been rarely discussed in the literature. However, people computing a Shapley value on each output dimension one-by-one can be encounter. To be more precise, one may first found natural to compute a Shapley value on the logit of the probability for each classes resuilting in a $N$-dimensional vector of the Shapley values for a $N$-classes problem. Such vector will be refered here as \emph{Shapley vector}. Even if the efficiency property holds, i.e. the sum of the element-wise logit of the base distribution with the Shapley vectors for each feature is equal to the element-wise logit of the prediction, the path from the base to the prediction may go out of the simplex which is conterintuitive. Indeed, this approach ignore the compositional nature of discrete probability distribution output by a classifier. Moreover, such strategy would require to run $N$ \emph{explanation model} contrary to our approach wich require a single explanation model.

As far as we know, this paper is the first work proposing an extension of the Shapley value framework to the multidimensional simplex for explaining probabilistic predictions in machine learning. We saw how the formalization of the standard Shapley value naturally extend to the simplex using the Aitchison geometry. The Shapley value therefore becomes a Shapley composition or distribution, i.e. a vector living on the probability simplex. Such Shapley composition informs on the contribution of a feature on the prediction. It tells how the feature moves the distribution from the base one to the predicted one on the simplex. The Aitchison geometry defines on the simplex an Euclidean vector space structure such that the Shapley compositions can be analysed through angles, norms and projections.

@@

\newpage
\section{Class-compositions}

A $k$-class-compositions $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class. This is a discrete probability distribution with maximum probability for the $k$th class and uniform values for the others. The $i$th part of the $\bm{c}^{(k)}$ is:
\begin{equation}
    c_i^{(k)} = \left\lbrace
  \begin{array}{@{}ll@{}}
    1-(N-1)p, & \text{if}\ i=k \\
    p, & \text{otherwise,}
  \end{array}\right.
\end{equation}
where $p<\frac{1}{N}$. We want the Aitchison norm of each class-composition to be one:
\begin{equation}
  \begin{aligned}
    \forall k \in \{1, \dots N \},~~~~~~\lVert \bm{c}^{(k)} \rVert_a = 1 \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^N \sum_{j=1}^N \left( \log \frac{c_i^{(k)}}{c_j^{(k)}} \right)^2} = 1,\\
    \text{for clarity,}&\\\text{we drop the $(k)$ from the equations,}&\\
    \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^{N}\left( \left(N-1 \right) \left( \log \frac{c_i}{p} \right)^2 + \left( \log \frac{c_i}{1-(N-1)p} \right)^2 \right)} = 1,\\
    \iff& \sqrt{\frac{1}{2N} 2(N-1) \left( \log \frac{p}{1-(N-1)p} \right)^2} =1,\\
    \text{since $p<\frac{1}{N}$ and the norm should be positive:}&\\
    \iff& \sqrt{\frac{N-1}{N}}\log \frac{1-(N-1)p}{p} =1,\\
    \iff& p = \frac{\exp \left( -\sqrt{\frac{N}{N-1}} \right)}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )}.
  \end{aligned}
\end{equation}
To summarize, the $i$th part of a $k$-class-compositions $\bm{c}^{(k)} \in \mathcal{S}^N$ is given by:
\begin{equation}
    c_i^{(k)} =
  \frac{1}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )} \left( \left\lbrace \begin{array}{@{}ll@{}}
     1, & \text{if}\ i=k \\
     \exp \left( -\sqrt{\frac{N}{N-1}} \right), & \text{otherwise,}
  \end{array}\right.\right).
\end{equation}
In this way, $\bm{c}^{(k)}$ is going straight to the direction of class $k$ and uniformly against all the others.

@@

Let $\{\hat{\bm{\phi}}_i\}_{1\leq i \leq d}$ be the estimated Shapley compositions (given by the Algorithm \ref{alg:2} in our experiments). Let $\displaystyle \bm{s}_{err} = \bm{f}(\bm{x}) \ominus \bm{f}_0 \ominus \underset{i=1}{\overset{d}\bigoplus} \hat{\bm{\phi}}_i$, where $\bm{f}_0$ is the base distribution, be the error composition on the pertubation of all Shapley compositions, i.e. the error making the efficiency property unfulfilled. In order to respect the efficiency property, we want this error to be the neutral element of the perturbation, i.e. the ``zero'' in the sense of the Aitchison geometry: the uniform distribution. We could simply perturb each estimated Shapley compositions by $\frac{1}{d}\odot\bm{s}_{err}$ however this would move each Shapley composition by the same amount while we want to allow the compositions with a higher estimation variance (i.e. with a precision likely to be lower) to move more than the ones with a smaller variance (i.e. with a precision likely to be higher).

>>>>

Compositional data analysis \cite{aitchison1982,pawlowskymodeling} aims at treating data---known as compositions---living on a simplex. Compositional data analysis tools have been applied for instance on geological, and chemical data but also to discrete probability distributions \cite{egozcue2011evidence,egozcue2018evidence,noe2023representing}. This paper presents the use of the Aitchison geometry of the simplex, coming from the field of compositional data analysis, for extending the concept of Shapley value to the probability simplex, i.e.~the space of discrete probability distributions. This can be used to properly explain multiclass probabilistic prediction.

@@

\section{Explaining a prediction with Shapley compositions}
\label{sec:explain}

Given a probabilistic prediction $\bm{f}(\bm{x}) \in \mathcal{S}^N$, the Shapley composition $\bm{\phi}_{\bm{f},\bm{x},\text{Pr}}(i)$ describes the contribution of the $i$th feature on the prediction. The efficiency property shows how the probability distribution moves from the base distribution, i.e.~the expected prediction regardless of the current input, to the prediction $\bm{f}(\bm{x})$. In the standard Shapley formulation recalled in Section \ref{sec:shapley}, the prediction is one-dimensional such that the Shapley quantity is a scalar. In applications where there are more than two possible classes, the prediction is multidimensional such that the Shapley quantity is too. Both live in the same space: the probability simplex. In this section, we discuss how the set of Shapley compositions can be analysed to better understand the contribution and influence of each feature on the prediction.

@@

The Shapley compositions can be visualized in the Euclidean space isometric to the simplex thanks to the ILR transformation presented in Section \ref{sec:ilr}. As we will see, this space has the advantage of being intuitive since it is a standard real $(N-1)$-dimensional vector space.

@@

In a three-classes case, the space of the predictions is $2$-dimensional. We illustrate this example with the well-known Iris classification dataset consisting of a set of flowers described by 4 features: sepal length and width, and petal length and width. The aim of the classification task is to predict to which of the three species (setosa, versicolor and virginica) a flower belongs. In the present example, a Support Vector Machine (SVM) with a radial basis function (rbf) kernel and with pairwise coupling \cite{wu2003probability} is used as a probabilistic classifier. Figure \ref{fig:3classes} shows the explanation of the classifier prediction with one versicolor instance as input. Figure \ref{fig:3classesshap} shows the Shapley compositions in the ILR space and Figure \ref{fig:3classesshapsum} shows how they move the base distribution to the prediction. Having the highest norm, the petal length is the feature contributing the most to the prediction and moves the base distribution to the versicolor maximum probability decision region (maximum probability decision region boundaries are the dashed gray lines). Class-compositions are represented by color dashed vectors. A class-composition is defined as an unit norm composition going straight to the direction of one class and against all the others (see Appendix \ref{app:classcompo} for a formal definition). The Shapley composition for the petal length is orthogonal to the virginica class-composition, this suggests that this feature does not contribute to the predicted probability for this class. The Shapley composition for the petal width goes straight to the opposite direction of the setosa class-composition suggesting that this feature contributes to rejecting this class. The other Shapley compositions have a low norm suggesting that these features do not contribute to the prediction.
\begin{figure}
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplot.png}
    \caption{Shapley compositions in the ILR space.}
    \label{fig:3classesshap}
  \end{subfigure}

\vspace{0.5cm}
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=0.8\linewidth]{figures/3classes/ilrplotsum.png}
    \caption{The sum of the Shapley compositions in the ILR space from the base distribution to the prediction.}
    \label{fig:3classesshapsum}
  \end{subfigure}
  \caption{Shapley explaination in the ILR space for the classification of an Iris instance.}
  \label{fig:3classes}
\end{figure}

@@

In a four-classes example, the simplex is $3$-dimensional. We illustrate this with a simple handwritten digit recognition task\footnote{We use the scikit-learn's digits dataset \cite{pedregosa2011scikit}.}. It consists of classifying an $8\times8$ image as representing one of the digits among 0, 1, 2 and 3. Since they are 64 pixels, considering each pixel as a feature would correspond to 64 Shapley compositions. This would be hard to visualize and analyse. We therefore reduce the number of features to 6 using a principal component analysis for better clarity and conciseness. A SVM with a rbf kernel and pairwise coupling is again used as a probabilistic classifier. The same explanation analysis as before can be applied here but within a $3$-dimensional plot as illustrated in Figure \ref{fig:4classesshapsum}. To better understand how this space is divided into four regions---each representing the maximum probability region for one class---one can think about the shape of a methane molecule. The hydrogens correspond to the vertices and the carbon to the center of a tetrahedron i.e.~a $3$-dimensional simplex. The relative positions of the class-compositions in the ILR space are the same as the bonds between the carbon and hydrogen: the angles are $\approx 109.5^{\circ}$. In this example, the tested instance is a 0\footnote{More examples can be obtained from the notebooks released on the git repository: link not shared during the double-blind peer reviewing.}.
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{figures/4classes/ilrplotsum.png}
  \caption{Shapley explaination in the $3$-dimensional ILR space for a four classes digit recognition task. The Shapley compositions are summed in the ILR space from the base distribution to the prediction. The gray transparent walls markout the maximum probability decision regions.}
  \label{fig:4classesshapsum}
\end{figure}

@@

When more than three classes are involved, all the dimensions of the ILR space cannot be visualized at once. However, $2$ or $3$-dimensional subspaces can still be visualized. In order to select the ILR components to investigate, one needs to understand what they refer to. In this section, we briefly discuss the interpretation of the ILR components.% of the explanation of a classifier prediction with more than three classes.

A component of the ILR space can be interpreted as a \emph{balance}, i.e.~a log-ratio of two geometrical means of parts \cite{egozcue2003isometric,egozcue2005groups,pawlowskymodeling}: one giving the central values of the probabilities in one group of classes and one for another group of classes. Therefore, a balance is here comparing the weight of two groups of classes. The set of balances is built such that they are geometrically orthogonal meaning they provide nonredundant information\footnote{Not to be confused with statistical uncorrelation \cite{pawlowskymodeling}.}.

This can be illustrated by a sequential binary partition or bifurcation tree. Figure \ref{fig:trees} gives two examples: \ref{fig:bifurc1} shows the bifurcation tree corresponding to the basis obtained with the Gram-Schmidt procedure as in \cite{egozcue2003isometric} which is the one used in the examples of Figures \ref{fig:3classes} and \ref{fig:4classesshapsum}  with respectively $N=3$ and $N=4$. Each node of the tree is a balance i.e.~an ILR component. The first balance $\tilde{p}_1$ first compares the probability for class $1$ with the probability for class $2$. Each next balance then recursively compares the probability for the next class with the probabilities for the previous classes independently of all the others.

In some applications, one may be interested in particular comparisons of groups of classes not necessarily given by a basis in the form of Figure \ref{fig:bifurc1}. For instance, as in an example presented in \cite{egozcue2005groups}, if one wants to compare political parties or groups, it may be pertinent to have a balance comparing left and right-wing groups. However, there are sometimes no obvious relevant comparisons to study. For instance, in the handwritten digit recognition problem, one may want to compare odd with even numbers or prime with non-prime but being basically a shape recognition problem, and the shape of the numbers being independent of their arithmetic properties, these comparisons are not pertinent. Be that as it may, the choice of the basis must be left open, whether or not it is based on a relevant strategy.

Let's use the basis of Figure \ref{fig:bifurc2} for a 10-classes digit recognition task\footnote{In this example, the bifurcation tree is obtained with agglomerative clustering of classes by recursively merging a pair of classes based on the Mahalanobis distance in the classifier's output space, assuming that within a pair of classes, the class-conditional densities are logistic-normal \cite{aitchison1980} with the same covariance matrix.}. We comment, for conciseness, a single $2$-dimensional subspace\footnote{More visualisations can be obtaned from the shared python notebooks.}. Let's have a look, in Figure \ref{fig:moreclasses35}, at the third and fifth ILR dimensions ($\tilde{p}_3$ and $\tilde{p}_5$). It is like saying we are only interested in comparing the probability assigned for class 0 with the probability assigned for class 6, and in comparing the probability assigned for class 1 with the group of probabilities assigned for classes 7 and 8. $\tilde{p}_3$ depends only on the probability for the digits 0 and 6 and $\tilde{p}_5$ depends only on the probabilities for the digits 1, 7 and 8. Therefore, the class-compositions for the other digits have a zero projection within this subspace and are therefore discarded in Figure \ref{fig:moreclasses35}. The class-compositions for 0 and 6 are orthogonal to the class-compositions for classes 1, 7 and 8. Indeed, the set of classes making the balance $\tilde{p}_3$ and the set of classes making $\tilde{p}_5$ have no intersection. In contrast, in the example of Figure \ref{fig:3classes}, $\tilde{p}_1$ is comparing the probabilities for the class setosa with the probability for the class versicolor and $\tilde{p}_2$ is comparing the probabilities for the class virginica with the group of probabilities for setosa and versicolor. In this case, none of the class-compositions are orthogonal. In Figure \ref{fig:moreclasses35}, since $\tilde{p}_5$ is comparing 1 with the group of digits 7 and 8, the projection on this line of the class-compositions for 1 goes in an opposite direction than the one for the class-compositions for 7 and 8. The latter two are equal and half as long as the former. In this way, $\tilde{p}_5$ compares the probability for 1 with the group of probabilities for 7 and 8 with the same weight. In other words, in this subspace, the class-compositions for 7 and 8 are reweighted such that this group of two classes has the same weight as the group made of the single class 1.

@@

  \vspace{0.5cm}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \input{bifurctree1.tex}
      \caption{Bifurcation tree used in the $10$-classes digit recognition task discussed in Section \ref{sec:balances} and in Figure \ref{fig:moreclasses35}.}
      \label{fig:bifurc2}
    \end{subfigure}
  \caption{Two examples of bifurcation tree.}
  \label{fig:trees}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{figures/moreclasses/ilrplot35.png}
  \caption{The sum of the Shapley compositions from the base to the prediction in the ILR subspace made of $\tilde{p}_3$ and $\tilde{p}_5$ for a test instance from class 2. $\tilde{p}_3$ compares the probability assigned for class 0 with the probability assigned for class 6 and $\tilde{p}_5$ compares the probability assigned for class 1 with the group of probabilities assigned for class 7 and 8. The color dashed vectors represent the class-compositions with non-zero projection.}
  \label{fig:moreclasses35}
\end{figure}
\subsection{Angles, norms and projections}

@@

The use of standard Shapley value for explaining machine learning model with multidimensional probabilistic output has been rarely discussed in the literature. However, people computing a Shapley value on each output dimension one-by-one can be encounter. To be more precise, one may first found natural to compute a Shapley value on the logit of the probability for each classes resuilting in a $N$-dimensional vector of the Shapley values for a $N$-classes problem. Such vector will be refered here as \emph{Shapley vector}. Even if the efficiency property holds, i.e.~the sum of the element-wise logit of the base distribution with the Shapley vectors for each feature is equal to the element-wise logit of the prediction, the path from the base to the prediction may go out of the simplex which is conterintuitive. Indeed, this approach ignore the compositional nature of discrete probability distribution output by a classifier. Moreover, such strategy would require to run $N$ \emph{explanation model} contrary to our approach wich require a single explanation model.

As far as we know, this paper is the first work proposing an extension of the Shapley value framework to the multidimensional simplex for explaining probabilistic predictions in machine learning. We saw how the formalization of the standard Shapley value naturally extend to the simplex using the Aitchison geometry. The Shapley value therefore becomes a Shapley composition or distribution, i.e.~a vector living on the probability simplex. Such Shapley composition informs on the contribution of a feature on the prediction. It tells how the feature moves the distribution from the base one to the predicted one on the simplex. The Aitchison geometry defines on the simplex an Euclidean vector space structure such that the Shapley compositions can be analysed through angles, norms and projections.

@@

\newpage
\section{Class-compositions}
\label{app:classcompo}
A $k$-class-compositions $\bm{c}^{(k)}  \in \mathcal{S}^N$ is defined as an unit norm composition going straight to the direction of the $k$th class. This is a discrete probability distribution with maximum probability for the $k$th class and uniform values for the others. The $i$th part of the $\bm{c}^{(k)}$ is:
\begin{equation}
    c_i^{(k)} = \left\lbrace
  \begin{array}{@{}ll@{}}
    1-(N-1)p, & \text{if}\ i=k \\
    p, & \text{otherwise,}
  \end{array}\right.
\end{equation}
where $p<\frac{1}{N}$. We want the Aitchison norm of each class-composition to be one:
\begin{equation}
  \begin{aligned}
    \forall k \in \{1, \dots N \},~~~~~~\lVert \bm{c}^{(k)} \rVert_a = 1 \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^N \sum_{j=1}^N \left( \log \frac{c_i^{(k)}}{c_j^{(k)}} \right)^2} = 1,\\
    \text{for clarity,}&\\\text{we drop the $(k)$ from the equations,}&\\
    \iff& \sqrt{\frac{1}{2N} \sum_{i=1}^{N}\left( \left(N-1 \right) \left( \log \frac{c_i}{p} \right)^2 + \left( \log \frac{c_i}{1-(N-1)p} \right)^2 \right)} = 1,\\
    \iff& \sqrt{\frac{1}{2N} 2(N-1) \left( \log \frac{p}{1-(N-1)p} \right)^2} =1,\\
    \text{since $p<\frac{1}{N}$ and the norm should be positive:}&\\
    \iff& \sqrt{\frac{N-1}{N}}\log \frac{1-(N-1)p}{p} =1,\\
    \iff& p = \frac{\exp \left( -\sqrt{\frac{N}{N-1}} \right)}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )}.
  \end{aligned}
\end{equation}
To summarize, the $i$th part of a $k$-class-compositions $\bm{c}^{(k)} \in \mathcal{S}^N$ is given by:
\begin{equation}
    c_i^{(k)} =
  \frac{1}{1+ (N-1)\exp \left( -\sqrt{\frac{N}{N-1}} \right )} \left( \left\lbrace \begin{array}{@{}ll@{}}
     1, & \text{if}\ i=k \\
     \exp \left( -\sqrt{\frac{N}{N-1}} \right), & \text{otherwise,}
  \end{array}\right.\right).
\end{equation}
In this way, $\bm{c}^{(k)}$ is going straight to the direction of class $k$ and uniformly against all the others.

@@

Let $\{\hat{\bm{\phi}}_i\}_{1\leq i \leq d}$ be the estimated Shapley compositions (given by the Algorithm \ref{alg:2} in our experiments). Let $\displaystyle \bm{s}_{err} = \bm{f}(\bm{x}) \ominus \bm{f}_0 \ominus \underset{i=1}{\overset{d}\bigoplus} \hat{\bm{\phi}}_i$, where $\bm{f}_0$ is the base distribution, be the error composition on the pertubation of all Shapley compositions, i.e.~the error making the efficiency property unfulfilled. In order to respect the efficiency property, we want this error to be the neutral element of the perturbation, i.e.~the ``zero'' in the sense of the Aitchison geometry: the uniform distribution. We could simply perturb each estimated Shapley compositions by $\frac{1}{d}\odot\bm{s}_{err}$ however this would move each Shapley composition by the same amount while we want to allow the compositions with a higher estimation variance (i.e.~with a precision likely to be lower) to move more than the ones with a smaller variance (i.e.~with a precision likely to be higher).
